A Thee-Step Approach for Assessing Landscape Connectivity via Simulated Dispersal: African Wild Dog Case Study

### 2022-02-07 Chapter_1/04_Manuscript/Manuscript.tex
updates

\newpage
\begin{abstract}
%------------------------------------------------------------------------------
%  LONG VERSION
%------------------------------------------------------------------------------
Dispersal of individuals contributes to long-term population persistence, yet
requires a sufficient degree of landscape connectivity. To date, connectivity
has mainly been investigated using least-cost analysis and circuit theory, two
methods that make assumptions that are hardly applicable to dispersal.
Least-cost analysis assumes that animals move towards a known endpoint and are
knowledgeable about the most favorable route. Circuit theory presumes a complete
random walk and fails to incorporate directional persistence. While these
assumptions can be relaxed by simulating dispersal movements explicitly across
the landscape of interest, a unified approach for such simulations is lacking.

Here, we present a simple three-step approach to simulate dispersal movements
and to assess connectivity using empirical GPS movement data and environmental
habitat layers. In step one, we use integrated step-selection functions to fit a
mechanistic movement model describing habitat and movement preferences of
dispersing individuals. In step two, we apply the parameterized movement model
to simulate dispersal trajectories. In step three, we derive three complementary
connectivity maps: a heatmap that highlights frequently traversed areas, a
betweenness map that pinpoints dispersal corridors, and a map of inter-patch
connectivity that indicates the presence and intensity of functional links
between habitat patches. As a case study, we applied the approach to GPS data
from dispersing individuals of the endangered African wild dog (\textit{Lycaon
pictus}) inhabiting the Kavango-Zambezi Transfrontier Conservation Area
(KAZA-TFCA).

Results from the case study show that wild dogs preferably disperse with
directional persistence in the vicinity to water bodies, through areas with
little human influence and sparse woodland cover. Dispersal simulations and
subsequent connectivity maps revealed several dispersal hotspots and corridors
across the extent of the KAZA-TFCA. Connectivity between NPs inside the
KAZA-TFCA was good, yet few dispersers successfully moved from Zambia's NPs into
neighboring areas. By rendering interactions between movement preferences and
habitat preferences, our movement model substantially outperformed a model that
omitted such interactions.

Ultimately, we show that a simulation-based approach that leverages on
step-selection functions offers a simple yet powerful alternative to traditional
connectivity modeling techniques. Such an approach not only makes fewer
biologically unrealistic assumptions but also permits a more mechanistic
understanding of dispersal and landscape connectivity. Our approach is thus
useful for a variety of applications in ecological, evolutionary, and
conservation research.

@@

% Importance of Dispersal & Connectivity
\subsection{Importance of Connectivity \& Connectivity Models}
Dispersal of individuals is a vital process that allows species to maintain
genetic diversity \citep{Perrin.1999, Perrin.2000, Frankham.2002, Leigh.2012,
Baguette.2013}, rescue non-viable populations \citep{Brown.1977}, and colonize
unoccupied habitats \citep{Hanski.1999b, MacArthur.2001}. Importantly, the
ability to disperse depends on a sufficient degree of landscape connectivity
\citep{Fahrig.2003, Clobert.2012}, making the identification and protection of
dispersal corridors that promote connectivity a task of fundamental importance
\citep{Nathan.2008, Doerr.2011, Rudnick.2012}. Identifying dispersal corridors
not only necessitates a comprehensive understanding of the factors that limit
dispersal movements, but also an appropriate model to estimate connectivity
\citep{Baguette.2013, Vasudev.2015, Hofmann.2021}. To date, the most commonly
used models to assess connectivity are least-cost path analysis (LCPA;
\citealp{Adriaensen.2003}) and circuit theory (CT; \citealp{McRae.2006,
McRae.2008}). Unfortunately, both models rest on assumptions that appear
unsuitable for dispersers, calling for the development of alternative
approaches. One promising alternative is to assess landscape connectivity via
simulated dispersal trajectories generated from individual-based movement models
(IBMMs, \citep{Diniz.2019}).

% Issues with Both Methods
\subsection{Issues with Traditional Connectivity Models}
Traditional connectivity models make assumptions that are rarely met for
dispersers. LCPA, for instance, assumes that individuals move towards a
preconceived endpoint and choose a cost-minimizing route accordingly
\citep{Sawyer.2011, Abrahms.2017}. While this assumption may be justifiable for
migrating animals, it is unlikely to hold for dispersers, as dispersers
typically move across unfamiliar territory towards an unknown endpoint
\citep{Koen.2014, Cozzi.2020}. CT, on the contrary, posits that animals move
according to a random walk, entailing that autocorrelation between subsequent
movements cannot be rendered \citep{Diniz.2019}. For dispersers, however,
autocorrelated movements are regularly observed \citep{Cozzi.2020,
Hofmann.2021}, implying that dispersal trajectories are usually strongly
directional. Moreover, because both models require static permeability or
resistance surfaces as input, they are unable to reflect the temporal dimension
of dispersal, meaning that statements about the expected duration for moving
between habitat patches are impossible \citep{Martensen.2017, Diniz.2019}.

% What about IBMMs?
\subsection{What about IBMMs?}
The shortcomings inherent to LCPA and CT can be overcome by simulating dispersal
using IBMMs and by converting simulated trajectories into meaningful measures of
connectivity \citep{Diniz.2019}. In contrast to LCPA and CT, IBMMs allow to
explicitly simulate how individuals move across and interact with the
encountered landscape \citep{Kanagaraj.2013, Clark.2015, Allen.2016,
Hauenstein.2019, Zeller.2020}, as well as to render potential interactions
between movement behavior and habitat conditions \citep{Avgar.2016}. This
strictly shifts the focus from a structural to a more functional view on
connectivity \citep{Tischendorf.2000, Kanagaraj.2013, Hauenstein.2019}.
Furthermore, IBMMs generate movement sequentially, i.e. they generate a series
of steps, so that the temporal dimension of dispersal movements becomes explicit
and allows modeling autocorrelation between consecutive steps
\citep{Diniz.2019}. Finally, simulations from IBMMs do not enforce movement or
connections towards preconceived endpoints, thereby preventing biases arising
from misplaced endpoints. Despite these advantages, a unifying framework to
simulate dispersal and assess connectivity using IBMMs is lacking. Considering
the large number of subjective decisions entailed by IBMMs, an approach that
streamlines and unifies the application of dispersal simulations to assess
connectivity will, however, be critical to safeguard interspecfic comparability
and reproducability.

% Proposed Solution: Three-Step approach
\subsection{Proposed Solution: Three-Step Approach}
Here, we propose and exemplify a simple three-step approach for simulating
dispersal and assessing landscape connectivity (\Cref{GraphicalAbstract}). In
step one, we combine GPS movement data of dispersing individuals with relevant
habitat covariates to fit a mechanistic movement model using integrated step
selection functions (ISSFs, \citealp{Avgar.2016}). ISSFs allow inference on the
study species' habitat kernel (i.e. habitat preferences), its movement kernel
(i.e. movement preferences/capabilities), and potential interactions between the
two \citep{Avgar.2016, Fieberg.2021}. In step two, we simulate individual
dispersal trajectories using the parametrized movement model. Comparable
simulations have already been applied to estimate steady-state utilization
distributions of resident individuals \citep{Potts.2013, Signer.2017} and to
model landscape connectivity, yet disregarding interdependencies between habitat
and movement kernels \citep{Clark.2015, Zeller.2020}. Finally, in step three, we
convert simulated trajectories into three complementary connectivity maps, each
highlighting a different aspect of connectivity; (i) a heatmap revealing areas
that are frequently traversed by dispersers (e.g. \citealp{Hauenstein.2019,
Zeller.2020}), (ii) a betweenness-map delineating dispersal corridors and
bottlenecks (e.g. \citealp{BastilleRousseau.2018}), (iii) and a map of
inter-patch connectivity, depicting the presence and intensity of specific
connections, as well as the average dispersal duration required for the
realization of those connections (e.g. \citealp{Gustafson.1996,
Kanagaraj.2013}).

% Introduction of the study species
\subsection{Case Study}
We showcase the application of the proposed approach (\Cref{GraphicalAbstract})
using GPS movement data collected on dispersing individuals of the endangered
African wild dog (\textit{Lycaon pictus}). The African wild dog is a highly
mobile species whose population persistence heavily relies on the availability
of large, natural or semi-natural landscapes and a sufficient degree of
connectivity among remaining subpopulations. Once common throughout sub-Saharan
Africa, this species has disappeared from much of its historic range, largely
due to human persecution, habitat fragmentation, and disease outbreaks
\citep{Woodroffe.2012}. Wild dogs typically disperse in single-sex coalitions
\citep{McNutt.1996, Behr.2020} and are capable of dispersing several hundred
kilometers \citep{DaviesMostert.2012, Masenga.2016, Cozzi.2020}. Although
previous studies have investigated connectivity using LCPA \citep{Hofmann.2021}
or CT \citep{Brennan.2020}, a more comprehensive and mechanistic understanding
of connectivity is missing for this species (but see \citealp{Creel.2020}). With
fewer than 6,000 free-ranging wild dogs remaining in fragmented subpopulations
\citep{Woodroffe.2012}, reliable information on landscape connectivity is
essential for the conservation of this endangered carnivore. Here, we use GPS
data of 16 dispersing wild dogs originating from a free-ranging population in
northern Botswana to parametrize a mechanistic movement model, which we then
employed to simulate 80,000 dispersal trajectories across the landscape of the
world's largest transboundary conservation area, the Kavango-Zambezi
Transfrontier Conservation Area (KAZA-TFCA). We anticipated that simulations
based on our three-step approach would overcome several of the above highlighted
conceptual shortcomings of traditional connectivity models and provide a more
comprehensive view on dispersal behavior and landscape connectivity.

@@

\section{Methods}
\subsection{Study Area}
Our simulation of dispersal trajectories and assessment of connectivity spanned
across the entire Kavango-Zambezi Transfrontier Conservation Area (KAZA-TFCA),
an area of approximately 1.3 Mio. km\textsuperscript{2} (\Cref{StudyArea}a and
b). The KAZA-TFCA is the world's largest transboundary conservation area and
comprises parts of Angola, Botswana, Namibia, Zimbabwe, and Zambia, thus hosting
a rich diversity of landscapes, ranging from savannah to grassland and from dry
to moist woodland habitats. In its center lies the Okavango Delta, a dominant
hydro-geographical feature and the world's largest flood-pulsing inland delta.
Large portions of the KAZA-TFCA are formally protected in the form of national
parks (NPs) or other protected areas, yet a considerable portion of the
landscape is still human-dominated (e.g. roads, agricultural sites, and
settlements).

@@

\subsection{Data Collection and Preparation}
\subsubsection{GPS Data}
We collected GPS movement data on 16 dispersing coalitions of African wild dogs
(7 female and 9 male coalitions) between 2011 and 2019 from a free-ranging
population in northern Botswana (details on the data collection can be found in
\cite{Cozzi.2020} and \cite{Hofmann.2021}). To delineate periods of dispersal,
we determined the exact timing of emigration from the natal pack and settlement
in a new territory using direct field observations and through visual inspection
of the net squared displacement (NSD) metric. The NSD metric measures the
squared Euclidean distance of a GPS relocation to a reference point
\citep{Borger.2012}, which we set to the center of the territory of each
dispersers' natal pack. Thus, dispersal was deemed to have started once an
individual left its natal territory and ended once the NSD metric remained
constant, indicating settlement. Because behavior during dispersal is more
pertinent to landscape connectivity than behavior during residence
\citep{Elliot.2014, Abrahms.2017}, we only considered data collected during
dispersal for our analysis. During dispersal, GPS collars recorded a fix every 4
hours and regularly transmitted data over the Iridium satellite system. To
ensure regular time intervals between GPS fixes, we removed any fixes that were
not successfully obtained at the desired 4-hour schedule (allowing for a
tolerance of \( \pm \) 15 minutes). We then converted the fixes (n = 4'169) into
steps, where each step represented the straight-line movement between two
consecutive GPS fixes \citep{Turchin.1998}.

\subsubsection{Habitat Covariates}
We represented the physical landscape in our study area by the habitat
covariates \textit{water-cover, distance-to-water, woodland-cover,
shrub/grassland-cover, and human-influence}. To render the seasonal dynamics in
water-cover of major water bodies for the extent of the Okavango Delta, we
applied an algorithm that enabled us to obtain weekly updated raster-layers for
water-cover and distance-to-water from MODIS satellite imagery
\citep{Wolski.2017, Hofmann.2021}. This algorithm is now implemented in the
\textsf{floodmapr} package (available on GitHub;
\url{https://github.com/DavidDHofmann/floodmapr}). To ensure a consistent
resolution across habitat covariates, we coarsened or interpolated all layers to
a resolution of 250 m x 250 m. A detailed description of how we prepared each
habitat covariate is provided in \cite{Hofmann.2021}.

Besides habitat covariates, we computed movement metrics that we used as
movement covariates in the ISSF models \citep{Avgar.2016, Fieberg.2021}.
Specifically, we computed for each step the step length (\textsf{sl}), its
natural logarithm (\textsf{log(sl)}), and the cosine of the relative turning
angle (\textsf{cos(ta)}). Moreover, we created the binary variable
\textsf{LowActivity}, indicating whether a step was realized during periods of
low wild dog activity (09:00 to 17:00 local time) or high wild dog activity
(17:00 to 09:00 local time, following \citealp{Cozzi.2012}). We performed all
data preparations, spatial computations, and statistical analysis in R, version
3.6.6 \citep{R.2020}. Some helper functions were written in {\tt C++} and
imported into {\tt R} using the {\tt Rcpp} package \citep{Eddelbuettel.2011,
Eddelbuettel.2013}.

\subsection{Step 1 - Movement Model}
We used ISSFs \citep{Avgar.2016} to parametrize a mechanistic movement model for
dispersing wild dogs. More specifically, we paired each realized (i.e. observed)
step with 24 random steps, so that a realized step plus its 24 random steps
formed a 25-step-stratum that received a unique identifier. As suggested by
\cite{Avgar.2016}, we generated random steps by sampling random turning angles
from a uniform distribution (\(-\pi, +\pi\)) (which is equivalent to a von Mises
distribution with location and concentration parameters; \(\mu = \kappa = 0\))
and step lengths from a gamma distribution that was fitted to realized steps
(scale \(\theta\) = 6'308 and shape \(k\) = 0.37).

Along each realized and random step, we extracted values from the habitat
covariate layers using the {\tt velox} package \citep{Hunziker.2021} and we
computed averages of each covariate along the steps. We further calculated the
movement metrics \textsf{sl}, \textsf{log(sl)}, and \textsf{cos(ta)} for each
step. To facilitate model convergence, we standardized all continuous covariates
to a mean of zero and a standard deviation of one. Correlations among covariates
were low (\(|r| < 0.6\); \citealp{Latham.2011}), so we retained all of them for
modeling.

To contrast realized steps (scored 1) and random steps (scored 0), we assumed
that animals assigned a selection score \(w(x)\) to each step (\ref{EQ1};
\citealp{Fortin.2005}). \(w(x)\) depended on the step's associated covariates
(\(x_1, x_2, ..., x_n\)) and on the animal's preferences (i.e. relative
selection strengths; \citealp{Avgar.2017}) towards these covariates (\(\beta_1,
\beta_2, ..., \beta_n\)):

@@

\noindent To estimate preferences (i.e. the \(\beta\)-coefficients), we used
mixed effects conditional logistic regression analysis \citep{Muff.2020} that we
implemented using the r-package {\tt glmmTMB} \citep{Brooks.2017}. The method
introduced by \cite{Muff.2020} allows to model random slopes, yet requires to
fix the variance of the stratum specific intercept to a large value. Hence we
fixed the stratum specific intercept variance to an arbitrary high value of \(10
^ 6\). We used disperser identity to model random slopes for all covariates.

The covariate structure of the movement model was based on a habitat selection
model that was previously developed for dispersing wild dogs (hereafter referred
to as \textit{based model}, \citealp{Hofmann.2021}). In the base model, no
interactions among habitat covariates and movement covariates were considered.
Hence, we expanded the base model and allowed for interactions between all
movement covariates and habitat covariates, thus reflecting that movement
behavior may depend on habitat conditions (details in Appendix A1). To determine
the most parsimonious movement model among model candidates, we ran stepwise
forward model selection based on Akaike's Information Criterion (AIC,
\citealp{Burnham.2002}). Finally, we validated the predictive power of the most
parsimonious model using k-fold cross-validation for case-control studies as
described in \cite{Fortin.2009}. This validation proves a significant prediction
in case the Spearman rank correlation of predicted step-ranks and associated
frequencies under the movement model is significantly greater than under the
assumption of random preferences (details in Appendix A2).

\subsection{Step 2 - Dispersal Simulation}
We used the most parsimonious movement model to simulate individual dispersal
trajectories within the study area. The simulation of a dispersal trajectory
resembled an ``inverted'' ISSF and was set up as follows. (1) We defined a
random source point and assumed a random initial orientation of the simulated
animal. (2) Starting from the source point, we generated 25 random steps by
sampling turning angles from a uniform distribution (\(-\pi, +\pi\)) and step
lengths from our fitted gamma distribution. Similarly to the empirical data,
each random step represented a 4 hours straight-line movement. To prevent
unreasonably large steps, we restricted sampled step lengths to a maximum of 35
km (i.e. the farthest dispersal distance traveled within 4 hours in our data).
(3) Along each random step, we extracted and averaged values from the different
habitat covariate layers and calculated movement covariates (as explained under
Step 1). To ensure compatible scales with the fitted movement model, we
standardized extracted values using means and standard deviations from the
empirical data. (4) We applied the parametrized movement model to predict the
selection score \(w(x)\) for each step using \ref{EQ1} and we converted
predicted scores into probabilities using \ref{EQ2}. (5) We randomly sampled one
of the generated random steps based on assigned probabilities and determined the
animal's new position. We repeated steps (2) to (5) until 2,000 steps were
realized.

>>>>

\newpage
\begin{abstract}
%------------------------------------------------------------------------------
%  LONG VERSION
%------------------------------------------------------------------------------
Dispersal of individuals contributes to long-term population persistence, yet
requires a sufficient degree of landscape connectivity. To date, connectivity
has mainly been investigated using least-cost analysis and circuit theory, two
methods that make assumptions that are hardly applicable to dispersal.
Least-cost analysis assumes that animals move towards a known endpoint and are
knowledgeable about the most favorable route. Circuit theory presumes a complete
random walk and fails to incorporate directional persistence. While these
assumptions can be relaxed by simulating dispersal movements explicitly across
the landscape, a unified approach for such simulations is lacking.

Here, we present a simple three-step approach to simulate dispersal movements
and to assess connectivity using empirical GPS movement data and a set of
habitat covariates. In step one, integrated step-selection functions are used to
fit a mechanistic movement model describing habitat and movement preferences of
dispersing individuals. In step two, the parameterized model is used to simulate
dispersal trajectories across the study area. In step three, a set of three
complementary connectivity maps is derived: a heatmap that highlights frequently
traversed areas, a betweenness map that pinpoints dispersal corridors, and a map
of inter-patch connectivity that indicates the presence and intensity of
functional links between habitat patches. As a case study, we applied the
approach to GPS data collected on dispersing individuals of the endangered
African wild dog (\textit{Lycaon pictus}) inhabiting the Kavango-Zambezi
Transfrontier Conservation Area (KAZA-TFCA) in southern Africa.

Besides providing insights into the movement and habitat preferences of
dispersing individuals, our movement model also rendered how movement behavior
depended on habitat conditions. The model thereby substantially outperformed a
model that omitted such interactions. While we simulated a total of 80,000
dispersal trajectories, metrics of connectivity only marginally changed beyond
10,500 simulations, suggesting that fewer simulations would have sufficed. The
three connectivity maps derived from simulations revealed several dispersal
hotspots and corridors across the extent of our study area. Each map emphasized
a different aspect of landscape connectivity, thus highlighting their
complementary nature. Overall, our case study demonstrates that a
simulation-based approach utilizing step-selection functions offers a simple yet
powerful alternative to traditional connectivity modeling techniques. Our
approach not only makes fewer unrealistic assumptions about dispersal but also
permits a more mechanistic understanding of dispersal movements and landscape
connectivity. It is thus useful for a variety of applications in ecological,
evolutionary, and conservation research.

@@

% Importance of Dispersal & Connectivity
\subsection{Importance of Connectivity \& Connectivity Models}
Dispersal of individuals is a vital process that allows species to maintain
genetic diversity \citep{Perrin.1999, Perrin.2000, Frankham.2002, Leigh.2012,
Baguette.2013}, rescue non-viable populations \citep{Brown.1977}, and colonize
unoccupied habitats \citep{Hanski.1999b, MacArthur.2001}. However, the ability
to disperse depends on a sufficient degree of landscape connectivity
\citep{Fahrig.2003, Clobert.2012}, making the identification and protection of
dispersal corridors that promote connectivity a task of fundamental importance
\citep{Nathan.2008, Doerr.2011, Rudnick.2012}. Identifying dispersal corridors
not only necessitates a comprehensive understanding of the factors that limit
dispersal, but also an appropriate model to estimate connectivity
\citep{Baguette.2013, Vasudev.2015, Hofmann.2021}. To date, the most commonly
used models are least-cost path analysis (LCPA; \citealp{Adriaensen.2003}) and
circuit theory (CT; \citealp{McRae.2006, McRae.2008}). Unfortunately, both
models rest on assumptions that appear unsuitable for dispersers, thus calling
for the development of alternative approaches. One promising alternative is to
assess landscape connectivity via simulated dispersal trajectories generated
from individual-based movement models (IBMMs, \citealp{Diniz.2019}). However,
the large number of subjective modeling decisions required by IBMMs renders
among-system comparisons difficult.

% Issues with Both Methods
\subsection{Issues with Traditional Connectivity Models}
Traditional connectivity models make assumptions that are rarely met for
dispersers. LCPA, for instance, assumes that individuals move towards a
preconceived endpoint and choose a cost-minimizing route accordingly
\citep{Sawyer.2011, Abrahms.2017}. While this assumption may be justifiable for
migrating animals, it is unlikely to hold for dispersers, as dispersers
typically move across unfamiliar territory towards an unknown endpoint
\citep{Koen.2014, Cozzi.2020}. CT, on the contrary, posits that animals move
according to a random walk, entailing that autocorrelation between subsequent
movements cannot be rendered \citep{Diniz.2019}. For dispersers, however,
autocorrelated movements are regularly observed \citep{Cozzi.2020,
Hofmann.2021}, meaning that dispersal trajectories are usually strongly
directional. Moreover, because both models require static permeability or
resistance surfaces as input, they are unable to reflect the temporal dimension
of dispersal, such that statements about the expected duration for moving
between habitat patches are impossible \citep{Martensen.2017, Diniz.2019}.

% What about IBMMs?
\subsection{What about IBMMs?}
The shortcomings inherent to LCPA and CT can be overcome by simulating dispersal
using IBMMs and by converting simulated trajectories into meaningful measures of
connectivity \citep{Diniz.2019}. In contrast to LCPA and CT, IBMMs allow to
explicitly simulate how individuals move across and interact with the
encountered landscape \citep{Kanagaraj.2013, Clark.2015, Allen.2016,
Hauenstein.2019, Zeller.2020}, as well as to render potential interactions
between movement behavior and habitat conditions \citep{Avgar.2016}. This
strictly shifts the focus from a structural to a more functional view on
connectivity \citep{Tischendorf.2000, Kanagaraj.2013, Hauenstein.2019}.
Furthermore, IBMMs generate movement sequentially, i.e. they generate a series
of steps, so that the temporal dimension of dispersal becomes explicit and
allows modeling autocorrelation between consecutive steps \citep{Diniz.2019}.
Finally, simulations from IBMMs do not enforce movement or connections towards
preconceived endpoints, thereby preventing biases arising from misplaced
endpoints. Despite these advantages, a unifying framework to simulate dispersal
and assess connectivity using IBMMs is lacking. Considering the large number of
subjective decisions entailed by IBMMs, an approach that streamlines and unifies
the application of dispersal simulations to assess connectivity will, however,
be critical to safeguard interspecfic comparability and reproducability among
studies.

% Proposed Solution: Three-Step approach
\subsection{Proposed Solution: Three-Step Approach}
Here, we propose and exemplify a simple three-step approach for simulating
dispersal and assessing landscape connectivity (\Cref{GraphicalAbstract}). In
step one, GPS movement data of dispersing individuals is combined with habitat
covariates to fit a mechanistic movement model via integrated step-selection
functions (ISSFs, \citealp{Avgar.2016}). ISSFs allow inference on the study
species' habitat kernel (i.e. habitat preferences), its movement kernel (i.e.
movement preferences/capabilities), as well as potential interactions between
the two \citep{Avgar.2016, Fieberg.2021}. In step two, the parametrized movement
model is used to simulate individual dispersal trajectories starting from
randomly placed source points. Comparable simulations have already been applied
to estimate steady-state utilization distributions of resident individuals
\citep{Potts.2013, Signer.2017} and to model landscape connectivity, yet
disregarding interdependencies between habitat and movement kernels
\citep{Clark.2015, Zeller.2020}. Finally, in step three, the simulated
trajectories are converted into three complementary connectivity maps; (i) a
heatmap revealing areas that are frequently traversed by dispersers (e.g.
\citealp{Hauenstein.2019, Zeller.2020}), (ii) a betweenness-map delineating
dispersal corridors and bottlenecks (e.g. \citealp{BastilleRousseau.2018}),
(iii) and a map of inter-patch connectivity, depicting the presence and
intensity of functional links between habitat patches, as well as the average
dispersal duration required for the realization of those connections (e.g.
\citealp{Gustafson.1996, Kanagaraj.2013}).

% Here, we use GPS data of 16 dispersing wild dogs originating from a
% free-ranging population in northern Botswana to parametrize a mechanistic
% movement model, which we then employed to simulate 80,000 dispersal
% trajectories across the landscape of the world's largest transboundary
% conservation area, the Kavango-Zambezi Transfrontier Conservation Area
% (KAZA-TFCA).

% Introduction of the study species
\subsection{Case Study}
We showcase the application of the proposed approach (\Cref{GraphicalAbstract})
using GPS movement data collected on dispersing individuals of the endangered
African wild dog (\textit{Lycaon pictus}). The African wild dog is a highly
mobile species whose population persistence heavily relies on the availability
of large, natural or semi-natural landscapes and a sufficient degree of
connectivity among remaining subpopulations. Once common throughout sub-Saharan
Africa, this species has disappeared from much of its historic range, largely
due to human persecution, habitat fragmentation, and disease outbreaks
\citep{Woodroffe.2012}. Wild dogs typically disperse in single-sex coalitions
\citep{McNutt.1996, Behr.2020} and are capable of dispersing several hundred
kilometers \citep{DaviesMostert.2012, Masenga.2016, Cozzi.2020}. Although
previous studies have investigated connectivity for this species using LCPA
\citep{Hofmann.2021} or CT \citep{Brennan.2020}, a more comprehensive and
mechanistic understanding of dispersal and connectivity is missing (but see
\citealp{Creel.2020}). Nevertheless, with about 6,000 free-ranging wild dogs
remaining in fragmented subpopulations \citep{Woodroffe.2012}, reliable
information on dispersal behavior and landscape connectivity is essential for
the conservation of this endangered carnivore. We anticipated that a
connectivity assessment based on our three-step approach would overcome several
of the above highlighted conceptual shortcomings of traditional connectivity
models, while providing a more detailed view on movement behavior during
dispersal.

@@

\section{Methods}
\subsection{Case Study}
\subsubsection{GPS Data}
We applied the three step approach mentioned above to GPS movement data from 16
dispersing African wild dog coalitions (7 female and 9 male coalitions). This
data has been collected between 2011 and 2019 from a free-ranging wild dog
population in northern Botswana (details on the data collection can be found in
\cite{Cozzi.2020} and \cite{Hofmann.2021}). During dispersal, GPS collars
recorded a fix every 4 hours and regularly transmitted data over the Iridium
satellite system. To ensure regular time intervals between GPS fixes, we removed
any fixes that were not successfully obtained at the desired 4-hour schedule
(allowing for a tolerance of \( \pm \) 15 minutes). To prepare the data for
step-selection analysis, we converted the fixes (n = 4'169) into steps, where
each step represented the straight-line movement between two consecutive GPS
fixes \citep{Turchin.1998}.

% To delineate periods of dispersal, we determined the exact timing of emigration
% from the natal pack and settlement in a new territory using direct field
% observations and through visual inspection of the net squared displacement (NSD)
% metric. The NSD metric measures the squared Euclidean distance of a GPS
% relocation to a reference point \citep{Borger.2012}, which we set to the center
% of the territory of each dispersers' natal pack. Thus, dispersal was deemed to
% have started once an individual left its natal territory and ended once the NSD
% metric remained constant, indicating settlement. Because behavior during
% dispersal is more pertinent to landscape connectivity than behavior during
% residence \citep{Elliot.2014, Abrahms.2017}, we only considered data collected
% during dispersal for our analysis.

\subsubsection{Study Area}
Our simulation of dispersal trajectories and assessment of connectivity spanned
across the entire Kavango-Zambezi Transfrontier Conservation Area (KAZA-TFCA,
\Cref{StudyArea}a and b). The KAZA-TFCA is the world's largest transboundary
conservation area and comprises parts of Angola, Botswana, Namibia, Zimbabwe,
and Zambia, thus hosting a rich diversity of landscapes, ranging from savannah
to grassland and from dry to moist woodland habitats. In its center lies the
Okavango Delta, a dominant hydro-geographical feature and the world's largest
flood-pulsing inland delta. Large portions of the KAZA-TFCA are formally
protected in the form of national parks (NPs) or other protected areas, yet a
considerable portion of the landscape is still human-dominated (e.g. roads,
agricultural sites, and settlements). We used NPs as patches to calculate
inter-patch connectivity (cf. Section 2.4.3). The decision to only consider NPs
as patches was purely out of simplicity and does not imply that dispersal
between other protected areas does not occur.

@@

\subsubsection{Covariates}
We represented the physical landscape in our study area by the habitat
covariates \textsf{water-cover, distance-to-water, woodland-cover,
shrub/grassland-cover, and human-influence}. To render the seasonal dynamics in
water-cover of major water bodies for the extent of the Okavango Delta, we
applied an algorithm that enabled us to obtain weekly updated raster-layers for
\textsf{water-cover} and \textsf{distance-to-water} from MODIS satellite imagery
\citep{Wolski.2017, Hofmann.2021}. This algorithm is now implemented in the {\tt
floodmapr} package (available on GitHub;
\url{https://github.com/DavidDHofmann/floodmapr}). To ensure a consistent
resolution across habitat covariates, we coarsened or interpolated all layers to
a resolution of 250 m x 250 m. A detailed description of how we prepared each
habitat covariate is provided in \cite{Hofmann.2021}.

Besides habitat covariates, we computed movement metrics that we used as
movement covariates in the ISSF models \citep{Avgar.2016, Fieberg.2021}.
Specifically, we computed for each step the step length (\textsf{sl}), its
natural logarithm (\textsf{log(sl)}), and the cosine of the relative turning
angle (\textsf{cos(ta)}), which is a measure of directionality
\citep{Turchin.1998}. Because wild dog activity is low during the hot midday
hours \citep{Cozzi.2012}, we additionally created the variable
\textsf{LowActivity}, indicating whether a step was realized during periods of
low wild dog activity (09:00 to 17:00 local time) or high wild dog activity
(17:00 to 09:00 local time). We performed all data preparations, spatial
computations, and statistical analysis in {\tt R}, version 3.6.6 \citep{R.2020}.
Some helper functions were written in {\tt C++} and imported into {\tt R} using
the {\tt Rcpp} package \citep{Eddelbuettel.2011, Eddelbuettel.2013}.

\subsection{Step 1 - Movement Model}
We used ISSFs \citep{Avgar.2016} to parametrize a mechanistic movement model
starting from GPS movement data collected on dispersing individuals. More
specifically, we paired each realized (i.e. observed) step with a set of 24
randomly generated alternative steps. A realized step plus its 24 random steps
formed a 25-step-stratum and received a unique identifier. As suggested by
\cite{Avgar.2016}, we generated random steps by sampling random turning angles
from a uniform distribution (\(-\pi, +\pi\)) (which is equivalent to a von Mises
distribution with location and concentration parameters \(\mu = \kappa = 0\))
and step lengths from a gamma distribution that was fitted to realized steps
(scale \(\theta\) = 6'308 and shape \(k\) = 0.37).

Along each realized and random step, we extracted values from underlying habitat
covariate layers using the {\tt velox} package \citep{Hunziker.2021} and we
computed averages of each covariate along the steps. For each step, we further
calculated the movement metrics \textsf{sl}, \textsf{log(sl)}, and
\textsf{cos(ta)}. To facilitate model convergence, we standardized all
continuous covariates to a mean of zero and a standard deviation of one.
Correlations among covariates were low (\(|r| < 0.6\); \citealp{Latham.2011}),
so we retained all of them for modeling.

To contrast realized steps (scored 1) and random steps (scored 0), we assumed
that animals assigned a selection score \(w(x)\) to each step (\ref{EQ1};
\citealp{Fortin.2005}), where \(w(x)\) depended on the step's associated
covariates (\(x_1, x_2, ..., x_n\)) and on the animal's preferences (i.e.
relative selection strengths; \citealp{Avgar.2017}) towards these covariates
(\(\beta_1, \beta_2, ..., \beta_n\)):

@@

\noindent To estimate preferences (i.e. the \(\beta\)-coefficients), we used
mixed effects conditional logistic regression analysis \citep{Muff.2020},
implemented through the r-package {\tt glmmTMB} \citep{Brooks.2017}. The method
introduced by \cite{Muff.2020} allows to model random slopes, yet requires to
fix the variance of the stratum specific intercept to a large value. We
therefore fixed the variance of the stratum specific intercept to an arbitrary
high value of \(10 ^ 6\) and used disperser identity to model random slopes for
all covariates.

The structure of our movement model was based on a habitat selection model that
was previously developed for dispersing wild dogs (hereafter referred to as
\textit{base model}, \citealp{Hofmann.2021}). In the base model, no interactions
among habitat covariates and movement covariates were considered, so we here
expanded the model and allowed for interactions between all movement covariates
and habitat covariates, thus reflecting that movement behavior could depend on
habitat conditions (details in Appendix A1). To determine the most parsimonious
movement model among model candidates, we ran stepwise forward model selection
based on Akaike's Information Criterion (AIC, \citealp{Burnham.2002}). We
validated the predictive power of the most parsimonious model using k-fold
cross-validation for case-control studies as described in \cite{Fortin.2009}.
This validation proves a significant prediction in case the Spearman rank
correlation coefficient of predicted step-ranks and associated frequencies under
the movement model is significantly greater than under the assumption of random
preferences (details in Appendix A2).

\subsection{Step 2 - Dispersal Simulation}
We used the most parsimonious movement model to simulate 80'000 individual
dispersal trajectories across the study area. The simulation of a dispersal
trajectory resembled an ``inverted'' ISSF and was set up as follows. (1) We
defined a source point and assumed a random initial orientation of the simulated
animal. (2) Starting from the source point, we generated 25 random steps by
sampling turning angles from a uniform distribution (\(-\pi, +\pi\)) and step
lengths from our fitted gamma distribution. (3) Along each random step, we
extracted and averaged values from the different habitat covariate layers and
computed the movement metrics \textsf{sl}, \textsf{log(sl)}, and
\textsf{cos(ta)}. To ensure compatible scales with the fitted movement model, we
standardized covariate values using means and standard deviations from the
empirical data. (4) We applied the parametrized movement model to predict the
selection score \(w(x)\) for each step using \ref{EQ1} and we converted
predicted scores into probabilities using \ref{EQ2}. (5) We randomly sampled one
of the generated random steps based on assigned probabilities and determined the
animal's new position. We repeated steps (2) to (5) until 2,000 steps were
realized.

As source points for the simulations, we distributed 50,000 points at random
locations inside protected areas that were larger than the average home range of
resident wild dog packs (i.e. \(>\) 700 km\textsuperscript{2};
\citealp{Pomilia.2015}). We placed another 30,000 points randomly inside the
buffer zone, mimicking potential immigration into the study area (Figure S1).
